Testing the motor and cognitive foundations of Paleolithic social transmission

### 2021-08-15 manuscript/manuscript.Rmd
update the result section and reinstall the R to load rjava package.

## Stone Tool Making

After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour. Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks. Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching. In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure**) of the kind of stone tools (flakes) they are being asked to produce. They were told that their objective was to produce as many flakes as possible from the materials provided. This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals. Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.

### Raw Materials

Each participant was provided with 9 cores for use over the 2-hour experiment. These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production. Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length x Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true comparing the two experimental conditions (Taught vs. Untaught mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.

@@

### Experimental Conditions

In both conditions, three researchers were present to record activities and collect materials. Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1). Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred. However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later. The order of cores used and associated hammerstones were recorded for each participant during the experiment.

@@

## Lithic Analysis

All finished cores were weighed and measured (L, W, T). Delta weight was calculated as (Start weight-End weight)/Start weight. All detached pieces (DPs) were collected and weighed. We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019]. All DPs larger than 40mm in maximum dimension were photographed and measured. It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off. We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons. Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks. Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al. (2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000). By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.

@@

## Statistical Analyses

To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002]. Information-theoretic approaches provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection. We used the corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between goodness of fit (likelihood of the data given the model) and parsimony (number of parameters). The AICc consists of the log likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples). A lower AICc indicates a more generalizable model and we used it to compare and rank various possible models. Each analysis begins with a full model that includes all predictors of interest. All possible combinations of predictors are then fit, and the resulting models are ranked and weighted based on their AICc. The "best" model is chosen because it has the lowest AICc score.

Continuous predictors were centered such that zero represents the sample average, and units are standard deviations. The full model was fitted with the lm function in R 3.2.3, and the glmulti package [@Bartoń 2015] was used for multi-modal selection and model comparison.

@@

# Results

Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>). Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences. The purpose of this section is to determine whether training impacted subject flaking performance and if any of the subject's individual psychometric and motor-skill aptitudes predict flaking performance.

## Principal Component analyses

The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in flake sizes and shapes and lithic performance measures.

### Detached Piece size and shape

To better understand the relationship between flake shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted. Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.

The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set **(Table**). Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor. Factor 2's loadings track the increasing relationship between thickness, length, and flake width. As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters. Factor 3 tracks the relationship between flake proximal and distal width relative to thickness. As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base. Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip. We used these three flake shape coordinates to approximate flake size and shape in the project's flake performance factor analysis.

### Lithic flaking performance measures

To better understand the relationship between our various lithic performance measurements and to reduce these data dimensionality, we conducted a second principal component analysis examining the study's seven lithic performance measures (mass of flakes relative to flaked core mass, count of large flakes [\>40mm and 5g], core delta mass, three flake shape factors, and the total number of cores used). All of these measures except the total cores flaked were summarized for each core. The Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3950, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.

The analysis yielded two factors explaining a total of 52.3% of the variance for the entire set of variables. Factor 1 tracks flaking quantity due to high positive loadings on mass flakes/flaked mass, core delta mass, total cores used, and flake shape factor 2. Negative loadings on flake factor 1 (flake size) and 3 (basal/tip width shape) suggest that flaking quantity comes at the expense of producing lots of small and thick stone splinters/chunks. This first factor explained 28% of the variance. Higher factor 1 values reflect higher flaking quantities. The second factor covers 24% of the sample variance. Factor 2 measures one's ability to carefully flake cores due to high positive loadings on large flake count and lower or neutral loadings on mass flakes/flaked mass, core delta mass, and total cores used. The resulting flakes are larger, relatively thinner, and more typically shaped due to high positive loadings on flake factor 1 (size) and flake factor 2 (relative thickness) contrasted with a high negative loading on flake shape factor 3 (basal/tip width). Higher factor two values show increased quality flaking performance.

**Figure** shows the covariance between our two flake performance principal components. While there is no significant relationship between the two factors, and the two study groups do not show significantly different slopes (p = 0.3), there are never-the-less differences between the two groups. Trained individuals are able to increase flake quality while increasing flake quality, whereas untrained individuals increase their flaking quantity at the expense of flake quality.

## Relationships between Performance Measures

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig. Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. R2 = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. R2 = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. R2 = 0.7), but no relation in the Trained group (Fig. Xb). A qualitatively similar trend with respect to Quality vs. Quantity (Fig. Xc) did not achieve significance (F[1, 17] = 0.6, p 0.2).

@@

## Do trained, untrained, and expert knappers perform differently?

Here we compare our flaking outcomes (flake size/shape and flaking performance factors) between the trained and untrained groups to see if training in any way impacted their flaking performance. We added a performance comparison (individual flake and core knapping performance metrics) between these two novice groups and our expert knapper to test for differences at different points in the stone flaking skill spectrum.

**Table** summarizes the group level performance tests. The results show no significant differences in flaking performance between the trained and untrained groups measured by our two flaking performance factors. Trained and untrained individuals overall performed equally as well in terms of flaking quality and quantity. Three-way flake size and shape comparisons between our expert knapper and the two novice groups show significant differences in flake shape factor 2 (relative thickness), but with a very low (\<0.01) effect size. This difference is driven by the expert's lower overall relative flake thickness. Regardless of training, our experimental subjects produced flakes that were on average the same size and shape as those of the expert trainer. We did, however, find several significant differences in the three-way comparisons of our individual flaking performance measures. The expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and on average had significantly smaller cores (effect size = 0.27) (**Figure**). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample (although they were not significantly different from the untrained group) (**Figures-Core examples too**). These results show that core reduction intensity and large flake production track this experiment's greatest differences between expert and novice performance.

## Does performance change over time?

Here we use the relative order in which each subject flaked their cores to test for changes in flaking performance across the 2hr experiment. For these analyses, we calculated the relative percentage for each core relative to the total cores knapped by each subject. These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons. Flaking outcomes were tracked using the two flake performance factors (quality and quantity flaking). We added the nodule starting mass to track whether training/practice times impacted raw material selection.

**Table** shows no significant training effects across the two flaking performance measures either as grouped data or between individuals (**Figures)**. This result demonstrated that flaking outcomes did not change dramatically across the study interval. The one significant main training effect related to core starting mass (with a strong main training effect size = 0.25). On average, core starting masses increase as subjects flake more cores, showing that subjects started with smaller nodules first that were smaller and easier to hold/flake. As the experiment wore on, they were left with larger and more challenging nodules. The small main effect of training condition is driven by higher starting nodule masses in the untrained group at the beginning, half way through, and at the end of the experiment.

## Do individual differences in motor skill and psychometric measures predict flaking performance?

One of the experiment's primary goals was to test if measures of individual variation in motor skill and intelligence/motivation predict success in stone flaking. To address this goal, we built two multivariate models examining the relations between our various psychometric measures, subject's motor skill scores, and our two lithic performance factors (quantity flaking and quality flaking). These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.

We have already demonstrated that subjects selected progressively larger nodules throughout the experiment. It is important now to understand whether nodule variability had any impact on our flaking results. Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, R2 = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences. Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, R2 = 0.41). However, these two measures track complementary components of athleticism (strength vs. speed/accuracy tradeoffs) and so we decided to include both in the model selection process.

We considered all possible interactions between five individual difference measures, core size, training condition, and the two lithic flaking performance factors (quantity and quality flaking) (wherein each subject provides one data point). Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitt's score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations.

### Model 1: Individual differences and quantity flaking

The first full model examined variance in the quantity of flaking tracked by our first performance factor explaining increases in the number of cores used, the degree of reduction on each core, and large/relatively thick flake production. The full model was fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process). The complete model statement is as follows:

*Quantity flaking* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength

From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 36 (Average AIC = 52). This model comprised the following statement with three main and four interaction effects:

*Quantity flaking* $\sim$ Training condition + Highest n-back level + BEAST score + Grip strength + Fitt's score$\cdot$Raven's Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)

This model explains a statistically significant and substantial proportion of variance in quantity flaking (R2 = 0.84, F (8, 12) = 8.2, p \< 0.01, adj. R2 = 0.74). A model evaluation comparing training vs. test data subsets shows no evidence for overfitting (Difference in R2 between training and test models = 0.02). A model residuals normality test shows no significant differences with the normal distribution (p = 0.35) indicating that this relationship (as required) is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (whether variance for all observations in our data set are the same) (BP = 8.2, df = 7, p = 0.3).

\
**Table** presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. Significant increases in quantity flaking were found for subjects with training (Est. = 0.63 [0.15, 1.11]), higher n-back levels (Est. = 0.71 [0.29, 1.13]), grip strength (Est. = 0.66 [0.36, 0.95]), and BEAST scores (Est. = 0.41 [0.04, 0.78]) regardless of starting nodule sizes (**Figure**). This suggests that several independent factors tracking the effects of training, social information use, visuo-spatial working memory, and strength improve an individual's ability to flake in greater quantities.

The model produced two significant interactions between training condition and n-back level and BEAST scores. Subjects in the trained group with higher n-back levels show lower quantity flaking scores (Est. = -1.35 [-1.85, -0.84]) while subjects in the trained group with higher BEAST scores also show lower quantity flaking scores (Est. = -0.83 [-1.37, -0.28]) when all the other variables are held constant. At first glance, these results appear to contrast with expectations regarding the effects of visuo-spatial working memory and social information use on success in technological tasks. However, looking at the matter more holistically it is clear that untrained subjects produce flaking quantities in a different way that relies on visuo-spatial working memory rather than taught information. When teaching is provided, being aware of one's surroundings (i.e. copying from other novices) can have a negative effect on flaking quantity. The interaction between BEAST scores and training shows that social information use acts to level the effects of an individual's propensity to "muscle" through the flaking task (both groups converge on an average quantity flaking score with higher BEAST scores). However, the direction of this change is different depending on training with untrained and socially inclined individuals increasing flaking quantities in contrast to trained and socially inclined individuals decreasing flaking quantities.

### Model 2: Individual differences and quality flaking

Our second model examining variance in quality flaking follows the same complete model statement we used for the quantity flaking with six covariates. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 41). This model comprised the following statement with three main and four interaction effects:

*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt's score + Training condition$\cdot$Grip strength + Nodule mass (as control)

This model explains a statistically significant and substantial proportion of variance in careful flaking (R2 = 0.78, F (8, 12) = 5.5, p \< 0.01, adj. R2 = 0.64) in the absence of any main training effects. A model evaluation comparing training vs. test data subsets shows some evidence for overfitting (Difference in R2 between training and test models = 0.17), which is unfortunately unavoidable with our small sample sizes. A model residuals normality test shows no significant differences with the normal distribution (p = 0.13) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 3.4, df = 8, p = 0.9).

>>>>

## Stone Tool Making
After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour. Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks. Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching. In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure**) of the kind of stone tools (flakes) they are being asked to produce. They were told that their objective was to produce as many flakes as possible from the materials provided. This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals. Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.

### Raw Materials
Each participant was provided with 9 cores for use over the 2-hour experiment. These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production. Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true comparing the two experimental conditions (Taught vs. Untaught mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.

@@

### Experimental Conditions
In both conditions, three researchers were present to record activities and collect materials. Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1). Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred. However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later. The order of cores used and associated hammerstones were recorded for each participant during the experiment.

@@

## Lithic Analysis
All finished cores were weighed and measured (L, W, T). Delta weight was calculated as (Start weight-End weight)/Start weight. All detached pieces (DPs) were collected and weighed. We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019]. All DPs larger than 40mm in maximum dimension were photographed and measured. It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off. We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons. Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks. Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al. (2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000). By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.

@@

## Statistical Analyses
To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002]. Information-theoretic approaches provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection. We used the corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between goodness of fit (likelihood of the data given the model) and parsimony (number of parameters). The AICc consists of the log likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples). A lower AICc indicates a more generalizable model and we used it to compare and rank various possible models. Each analysis begins with a full model that includes all predictors of interest. All possible combinations of predictors are then fit, and the resulting models are ranked and weighted based on their AICc. The "best" model is chosen because it has the lowest AICc score.

Continuous predictors were centered such that zero represents the sample average, and units are standard deviations. The full model was fitted with the lm function in R 3.2.3, and the glmulti package was used for multi-modal selection and model comparison.

@@

# Results
Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>). Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences. We were particularly interested in: 1) group level effects of experimental condition, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences. To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject’s motor skill scores, and our two lithic performance factors.

## Principal Component analyses
The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.

### Detached Piece size and shape
To better understand the relationship between DP shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted. Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.

The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set **(Table**). Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor. Factor 2's loadings track the increasing relationship between thickness, length, and flake width. As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters. Factor 3 tracks the relationship between flake proximal and distal width relative to thickness. As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base. Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip. We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.

### Lithic flaking performance measures
To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study’s six lithic performance measures (count of large pieces [>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors). All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures. Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.

The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table). Factor 1 (hereafter “Quantity”) explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass. Performance factor 2  (hereafter “Quality”) covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, “flake-like” shape) and a negative loading on Shape Factor 2 (“splinter-like” thickness and elongation). High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.

These two factors address flaking performance at the level of individual cores, however we were also interested in the overall productivity/rate of work of each participant over the entire two hours. For example, looking at a knappers average Quality and Quantity factor scores would not differentiate between a participant who spent the entire time exhaustively reducing one core vs. another participant who did the same to all nine of their allotted cores in the same time. To capture this aspect of variation we calculated a simple Total Productivity metric as the sum of all mass a participant removed from cores during the experiment.

## Relationships between Performance Measures
This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig. Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the Trained group (Fig. Xb). A qualitatively similar trend with respect to Quality vs. Quantity (Fig. Xc) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

@@

## Do trained, untrained, and expert knappers perform differently?
Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups. Our expert demonstrator/instructor is included as a performance benchmark.

Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (Figure). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group’s data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too). We also observed a significant difference in shape factor 2 (splinters) driven by the expert’s lower values, but with a very low (<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.

While we did not find significant differences in central tendency between our two experimental groups, results (Figure and Figure) did indicate lower variance in the trained group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances. Table and Figure present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass). In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (Figure), rather than to shift the mean.

## Does performance change over time?
In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.  For these analyses, we calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8 both equal 50% complete). These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons. Flaking outcomes were tracked using the two performance factors (Quality and Quantity). We added the nodule starting mass to track whether training/practice times impacted raw material selection.

Table shows no significant training effects across the two performance measures either as grouped data or between individuals (Figures). This result demonstrated that flaking outcomes did not change dramatically across the study interval. This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure). The one significant main training effect related to core starting mass (with a strong effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure).

## Do individual differences in motor skill and psychometric measures predict flaking performance?
One of the experiment’s primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions. To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality). These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.

We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point. Each model’s continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations. Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41). However, these two measures track complementary components of athleticism (strength vs. speed/accuracy tradeoffs) and so we decided to include both in the model selection process.

The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package’s automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process). All three models follow the same complete model statement as follows:

*Flaking performance variable $\sim$ Training condition + Highest n-back level + Raven’s Progressive Matrix score + BEAST score + Fitt’s score + Grip strength*

For our two per-core performance factors (Quantity and Quality) it is also relevant to consider how individual core features may have affected performance. We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models. We did, however, find that subjects selected progressively larger nodules throughout the experiment. It is thus important to understand whether nodule variability had any impact on our flaking results. Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p < 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.

### Model 1: Individual differences and quantity flaking
Our first model examined variance in overall flaking productivity measured by each subject’s combined flaked mass (nodule starting mass-core final mass). This provides a basic measure of variation in individuals’ success detaching pieces and reducing cores from a standardized (see Methods) raw material supply. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13). This model comprised the following statement with two main and three interaction effects:

*Total flaked mass $\sim$ Training condition + Grip strength + RPM$\times$Highest n-back level + Fitts score$\times$BEAST score + Grip strength$\times$RPM*

This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p < 0.01, adj. $R^2$ = 0.77). A model residuals normality test shows no significant differences with the normal distribution (p = 0.72) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).

Table presents this model’s coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p< 0.05) and substantial (Standardized Estimate >= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (Figure), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.

Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (Figures). In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group. Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition (Coolidge and Wynn 2005) and social learning (Morgan et al. 2015) in the acquisition of knapping skills. Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.

### Model 2: Individual differences and quality flaking
The second full model examined the variance in average flaking Quantity per core. It thus complements our first model assessing overall productivity by testing for differences in reduction intensity at the level of individual cores. From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 44). This model comprised the following statement with three main and four interaction effects:

*Quantity $\sim$ Highest n-back level  + BEAST score + Fitt’s score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)*

This model explains a statistically significant and substantial proportion of variance in quantity flaking ($R^2$ = 0.7, F (8, 12) = 3.6, p = 0.02, adj. $R^2$ = 0.5). A model residuals normality test shows no significant differences with the normal distribution (p = 0.38) indicating that this relationship (as required) is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (whether variance for all observations in our data set are the same) (BP = 4.4, df = 8, p = 0.8).

The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength. As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (Figure).

Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (Figure). Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.

### Model 3: Individual differences and quality flaking
Our third model examining variance in Quality follows the same complete model statement we used for Quantity. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 39). This model comprised the following statement with three main and four interaction effects:

*Quality flaking $\sim$ Highest n-back level  + Fitt’s score + Grip strength + Fitt’s score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt’s score + Training condition$\times$Grip strength + Nodule mass (as control)*
